Historical trends in the contribution of six major causes of death to the black-white life expectancy gap by US state

### 2017-04-10 Code/combined-meta-regression.Rmd

```{r load-data}
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
cod_decomp_results$COD_cont_yrs_SE <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl)/3.92
```

@@

```{r save_data}
write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
```

>>>>

```{r load-data}
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
cod_decomp_results$COD_cont_yrs_SE <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl)/3.92
```

@@

```{r save_data}
write.csv(cod_decomp_results, "/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
```

### 2017-04-10 Manuscript/Manuscript_COD_Trends.Rmd

```{r load_data, echo=F}
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")

cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
```

@@

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.

@@

The over-arching objective of this study was to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females. This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states. Ideally, this study may facilitate the identification of one policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.

@@

To describe the general pattern in these trend lines we used random effects meta regression. Each state's mean contribution was regressed against a linear spline for calendar time that was estimated with five degrees of freedom. The model included random effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year. To identify divergent states, we compared the state-specific trends (estimated separately for each state using the Bayesian model) to the general pattern estimated using meta regression.

@@

**Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years. The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends. These lines are overlaid with the meta-estimated common trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).

@@

The situation was reversed for females, with black females experiencing relatively larger mortality decreases over time compared to white females. Thus, the contribution of CVD to the life expectancy gap decreased by more than a year in several states (**Figure 4**). Of note, some states experienced a period of early increasing contributions, most notably, Arizona, Washington, and Oregon, that was not captured by the general pattern. Furthermore, Wisconsin and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.

@@

For all races and sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**). Among males, this inverted U-shape pattern was accentuated among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**). Notable exceptions to this general pattern include Washington DC and Mississippi, states that underwent increasing cancer contributions over time, and New Mexico and Rhode Island, exhibiting decreasing contributions over time (although imprecisely estimated).

For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern is similar between blacks and whites and most states do not exhibit the accentuated period of increase in mortality that was estimated for black males. Thus, the contribution of cancer to the life expectancy gap is generally flat over time (**Figure 6**). [You could potentially stop here to save space and just include the comment about there being little inequality in cancer mortality ]. Washington DC deviates from this pattern, and exhibited increasing cancer contributions to the gap, while Rhode Island exhibited the reverse. Most notably, some states exhibited no inequality in cancer mortality (**eFigures 3 and 4**), including New York, and Connecticut (and for males, these states recently eliminated the black-white difference in cancer mortality). Furthermore, Massachusetts and (imprecisely) Rhode Island exhibit lower cancer rates among black women and men compared to whites in 2013.

@@

Limitations
-does not account for migration. Washington DC - have a lot of migration of high SES white folks coming in, and perhaps also relatively high SES black folks moving out
-does not account for hispanic ethinicity - demographic shifts in FL for example of black and white hispanics which will pull LE and disparities

>>>>

```{r load_data, echo=F}
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")

cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
```

@@

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethinicity. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.

@@

The over-arching objective of this study is to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females. This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states. This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.

@@

To describe the general pattern in these trend lines we used random effects meta regression. Each state's mean contribution was regressed against a linear spline for calendar time that was estimated using four internal knots. The model included random effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year. To identify divergent states, we compared the state-specific trends (estimated separately for each state using the Bayesian model) to the general pattern estimated using meta regression.

@@

**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for males and females, in years. The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends. These lines are overlaid with the meta-estimated common trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).

@@

The situation was reversed for females, with black females experiencing relatively larger mortality decreases over time compared to white females. Thus, the contribution of CVD to the life expectancy gap decreased by more than a year in several states (**Figure 4**). Of note, some western states experienced a period of early increasing contributions that was not captured by the general pattern. Furthermore, Wisconsin, Michigan, and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.

@@

For both races and both sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**). Among males, this inverted U-shape pattern was accentuated among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**). Notable exceptions to this general pattern include Washington DC and Wisconsin, states that had higher initial contributions and underwent increasing cancer contributions over time, and New York and Massechusetts, exhibiting sharper decreases in cancer contributions.

For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**). [You could potentially stop here to save space and just include the comment about there being little inequality in cancer mortality ]. Washington DC deviates by exhibiting increasing cancer contributions to the gap, while New York, Connecticut, exhibit no inequality in cancer mortality, while Massachusetts exhibits lower cancer rates among blacks (**eFigure 4**).

@@

Lit review

Life Expectancy generally
The spotlight on racial disparities in the US has intensified. Several recent papers examining US mortality by region, cause of death, or within population subgroups have received considerable attention while ignoring important black-white disparities.2, 3, 4 These disparities remain profound, and no previous paper has examined cause-specific contributions by state over an extended time period. Our analysis provides an important contribution to the literature by highlighting a high degree of heterogeneity in state-specific performance over time in eliminating these inequalities.
from above:
2. Chetty R, Stepner M, Abraham S, Lin S, Scuderi B, Turner N, Bergeron A, Cutler D. The association between income and life expectancy in the United States, 2001-2014. JAMA. 2016. 315(16):1750-1766.

3. Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al. US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014. JAMA. 2016. 316(22):2385-2401.

4. Case A, Deaton A. Rising morbidity and mortality in midlife among white non-Hispanic Americans in the 21st century. Proc Natl Acad Sci. 2015. 112(49):15078-83.

Cancer

Oliver. CEBP. 2008. Recent trends in BW disparities in Cancer Mortality. Note: higher mort rates in their paper likely because they restrict to >35yo.
-BW diff in lung cancer mort mimics trends in overall mort rate in males. In females, no diff in lung cancer (although minor difference in other cancer related to smoking)
-increasing disparity is seen in colorectal cancer in males between B and W
-hump also seen in prostate cancer, especially among blacks, but falling mort rates since mid 1990s.
-among females, increasing disparity in mort rates from colorectal cancer and breast cancer since 1975 (beginning of their study)
-not sure what explains the inverted-U in cancer mort in our data among females - this CEBP people doesn't seem to shed light on this.

Screening - when is it proven effective vs. when does it affect the recorded COD? Corinne to review that literature.
-breast cancer: kills mostly among those 70+ yo.

-trends in incidence may be related to overdx (prostate cancer, for ex) - for prostate cancer don't know whether the decline is due to improvements in treatment or due to screening -- could be both or just one

https://nccd.cdc.gov/uscs/toptencancers.aspx#White
-black women have higher rates of breast cancer than white women (28/100000 vs 20 for mortality)
-blacks have higher rates of prostate cancer than whites (39.1 vs 18)

Smoking and lung cancer
see: https://www.washingtonpost.com/news/to-your-health/wp/2015/11/12/smoking-among-u-s-adults-has-fallen-to-historic-lows-these-7-charts-show-who-still-lights-up-the-most/?utm_term=.f2b261e63acf -- uses CDC data for citation
-smoking rates have fallen steadily over the course of the study time period, but homogeneity in smoking still exists. used to be 42.4% in 1965 and down to 16.8% in 2014.
-by race: nearly same for blacks and whites - 17.5% in blacks vs. 18.2% in whites, but 11.2% in hispanics (which are mixed into BW designation in our study)
-by geography: highest in the midwest (20.7%), then south (17%), NE (15.3%), then west (13%), within heterogeneity within state (
Dwyer-Lindgren  L, Mokdad  AH, Srebotnjak  T, Flaxman  AD, Hansen  GM, Murray  CJ.  Cigarette smoking prevalence in US counties: 1996-2012.  Popul Health Metr. 2014;12(1):5.)

Obesity and...
-high obesity in the south - ref 9
-women and blacks have highest rates of obesity - ref 8

Opiods
https://www.cdc.gov/nchs/products/databriefs/db273.htm - Although rates increased 63% for non-Hispanic black persons (from 7.5 per 100,000 in 1999 to 12.2 in 2015) and 43% for Hispanic persons (from 5.4 in 1999 to 7.7 in 2015), non-Hispanic white persons experienced the greatest increase between 1999 and 2015. The age-adjusted rate of drug overdose deaths among non-Hispanic white persons increased 240% from 6.2 per 100,000 in 1999 to 21.1 in 2015.

Centers for Disease Control and Prevention. Vital signs: overdoses of prescription opioid pain relievers and other drugs among women—United States, 1999–2010. MMWR Morb Mortal Wkly Rep. 2013;62(26):537–542: opioid-related mortality increased 415% among women and 265% among men between 1999 and 2010.38

Paulozzi LJ, Xi Y. Recent changes in drug poisoning mortality in the United States by urban-rural status and by drug type. Pharmacoepidemiol Drug Saf. 2008;17(10):997–1005.: in 1999, large central metropolitan areas had the highest opioid-related mortality rates, and noncore areas had the lowest rates; by 2004, noncore areas had the highest rates and had seen the largest relative increase during that time period.50

Paulozzi LJ, Ryan GW. Opioid analgesics and rates of fatal drug poisoning in the United States. Am J Prev Med. 2006;31(6):506–511: drug poisoning mortality rates varied 8-fold between states, as did opioid consumption (methadone, 13-fold; oxycodone, 7-fold; overall, 4-fold).17

Centers for Disease Control and Prevention. Vital signs: overdoses of prescription opioid pain relievers—United States, 1999–2008. MMWR Morb Mortal Wkly Rep. 2011;60(43):1487–1492.: Another study showed that prescription opioid sales ranged from 3.7 (Illinois) to 12.6 (Florida) kilograms per 10,000 population, and opioid-related mortality rates ranged from 5.5 (Nebraska) to 27 (New Mexico) per 100,000 population.2

Strengths

Limitations
-does not account for migration. Washington DC - have a lot of migration of high SES white folks coming in, and perhaps also relatively high SES black folks moving out
-see the migration literature sent by Sam
-does not account for hispanic ethinicity - demographic shifts in FL for example of black and white hispanics which will pull LE and disparities
